# M10 Notes

## R vs Python

-   Indexing: $0$ vs $1$ based.
-   Meaning of negative numbers in slices, e.g. `foo[-1]`.
-   Dots don't mean anything in R.
-   Assignment operator in R is `<-` or `->`.
    -   Can use `=` but be careful ... Can be confused with argument assignment in functions.

## R Data Structures

| **Structure** | **Dim** | **Data Type** | **Shape** | **Python**        |
|---------------|---------|---------------|-----------|-------------------|
| Vector        | 1       | uniform       | sequence  |                   |
| Matrix        | 2       | uniform       | square    |                   |
| Array         | 𝑁       | uniform       | cube+     | NumPy array       |
| List          | 1       | non-uniform   | ragged    | List, Dict        |
| Data Frame    | 2       | multiple      | uniform   | Pandas Data Frame |

-   These reflect the evolution of R.
-   We mainly use Vectors and Data Frames.

## Factors

-   A kind of data structure, since they organize data types.
    -   They can store both strings and integers.
-   The categorize data as *levels*.
    -   Levels are unique, like sets in Python or LOVs in SQL.
    -   Levels are stored alongside the vector.
    -   Levels constrain what can be added to the factor vector.
    -   Levels are always characters, even when the data are numeric or boolean
-   Useful in data frame columns have list of values that make sense as a group.
    -   Male, Female
    -   True, False
    -   1, 2, 3, 4, 5
-   Created with `factor()` taking a vector as input.
-   Analogous to Categories in Pandas.

**Example**

```{r}
v1 = c(1,5,6,9,4,3,5,8,7,6,3,0,0,0,1,2,3,6,4,5,7,9)
```

```{r}
levels(v1)
```

```{r}
f1 = factor(v1)
```

```{r}
levels(f1)
```

```{r}
print(f1)
```

```{r}
nlevels(f1)
```

# Levels act as a constraint on the factor vector.

```{r}
f1[5]
```

```{r}
f1[5] <- 20
```

```{r}
f1
```
