---
editor: 
  markdown: 
    wrap: 72
---

# NB: Control Structures and Functions

## `if`/`else` Statements

You've seen this in Python.

They work the same way in both languages.

Here's their syntax in R.

```         
if(<condition>) {
        ## do something
} 

if(<condition>) {
        ## do something
} 
else {
        ## do something else
}

if(<condition1>) {
        ## do something
} else if(<condition2>)  {
        ## do something different
} else {
        ## do something different
}
```

Generate a uniform random number:

```{r}
x <- runif(1, 0, 10) # From the Uniform Distribution

if(x > 3) {
    y <- 10
} else {
    y <- 0
}

x
y
```

You assign an `if` statement to a variable.

```{r}
z <- if(x > 3) {
  10
} else { 
  0
}

z
```

You can stack `if` blocks, too.

```         
if(<condition1>) {

}

if(<condition2>) {

}
```

## Control Structures

### `for` Loops

For loops are straight-forward. The take an interator variable, e.g.
`i`, and assign it successive values from a sequence or vector.

For loops are often used to iteratE over the elements of an object
(list, vector, etc.).

```{r}
for(i in 1:10) {
  print(i)
}
```

According to Hadley Wickham, loops are pretty much the only looping
construct that you will need in R.

The following three loops all have the same behavior.

```{r}
x <- c("a", "b", "c", "d")
for (i in 1:4) {
  # Print out each element of 'x'
  print(x[i])  
}
```

### `seq_along()`

The `seq_along()` function is commonly used in conjunction with for
loops in order to generate an integer sequence based on the length of an
object (in this case, the object x).

```{r}
x
```

Generate a sequence based on length of 'x':

```{r}
for(i in seq_along(x)) {   
  print(x[i])
}
```

It is not necessary to use an index-type variable.

```{r}
for(letter in x) {
  print(letter)
}
```

For one line loops, the curly braces are not strictly necessary.

```{r}
for(i in 1:4) print(x[i])
```

### Nested `for` loops

Dor loops can be nested inside of each other.

```{r}
x <- matrix(1:6, 2, 3)
for(i in seq_len(nrow(x))) {
  for(j in seq_len(ncol(x))) {
    print(x[i, j])
  }   
}
```

Nested loops are used to generate multidimensional or hierarchical data
structures (e.g. matrices, lists).

## `while` Loops

As with Python, while loops start with a condition. It loops while the
condition is true and stops when it is false.

Remembe, while loops can go on forever is the truth condition is never
met.

```{r}
count <- 0
while(count < 10) {
  print(count)
  count <- count + 1
}
```

## `repeat` Loops

`repeat` loops are using by R. They initiate an infinite loop right from
the start.

The only way to exit a repeat loop is to call break on an internal
condition.

```{r eval=FALSE, include=FALSE}
x0 <- 1
tol <- 1e-8

repeat {
  x1 <- computeEstimate()
  if (abs(x1 - x0) < tol) {  ## Close enough?
    break
  } else {
    x0 <- x1
  } 
}
```

## `next` and `break`

next is used to skip an iteration of a loop. Same as Python `continue`.

```{r}
for (i in 1:100) {
  if (i <= 20) {
    # Skip the first 20 iterations
    next                
  }      # Do something here
}
```

`break` is used to exit a loop immediately.

```{r}
for (i in 1:100) {
  print(i)
  if (i > 20) {
    # Stop loop after 20 iterations
    break  
  }     
}
```

## Functions

Define some data

```{r}
x <- 5                
xx <- c(4, 6, 7, 8, 2, 11)
```

Now, define a function that does the following: - takes a value, vector
of values as inputs - normalizes the value against the vector by
subtracting the vector mean from value, and dividing by vector standard
deviation.

```{r}
compute_zscore <- function(val, vec) {
  z <- (val - mean(vec)) / sd(vec)
}

print(compute_zscore(x, xx))
```

If vector contains identical values, `sd` is zero, and so the z-score is
undefined.

```{r}
print(compute_zscore(x, c(1, 1, 1, 1)))
```

If vector contains missing values, the result will be `NA`.

```{r}
xx_na <- c(1, NA, 3, 5) 
print(compute_zscore(x, xx_na))
```

## Using conditions in functions

A function returns 1 if passed value is odd, 0 if even

## %% is mod operator (returns remainder)

```{r}
is_odd <- function(x){ if (x %% 2 == 1) { 
    return(1) 
  } else { 
    return(0)
  } 
}
```

Call to test some cases:

```{r}
is_odd(4)
is_odd(3)
```

Function arguments can use default values:

```{r}
threshold_vals <- function(p, thresh = 0.5) {
  # for each element in p, returns TRUE if value > thresh, else FALSE
  p > thresh
}
```

```{r}
threshold_vals(c(0.6, 0.4, 0.1, 1))
```

Now, pass a threshhold:

```{r}
threshold_vals(c(0.6, 0.4, 0.1, 1), 0.7)
```

Assert important preconditions

```{r}
add_vectors <- function(x, y) {
  # assert the lengths of vectors x and y match
  # if they do, sum elementwise, else throw error with stop()

  if (length(x) != length(y)) {
    stop("x and y must be the same length", call. = FALSE)
  }
  x + y
}
```

```{r}
add_vectors(c(1, 2, 3), c(3, 3, 3))
```

```{r}
#| eval: false
add_vectors(c(1, 2, 3), c(3, 3, 3, 3)) # breaks
```

## Scoping Rules

```{r}
z <- 4
test_fcn <- function(x) {
  x^z
}
```

Now look at this:

```{r}
test_fcn(2)
```

If `z` isn't defined in the function, how does this work?

R's scoping rules are similar to Python's.

Since `z` isn't in the function, **R looks in the function's
environment** for it.

For more on scoping rules, see [Chapter 15: Scoping Rules of
R](https://bookdown.org/rdpeng/rprogdatascience/scoping-rules-of-r.html)
in Peng's *R Programming For Data Science*.